Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-07-24 latex/vanishing.tex
Compilesgit statusgit status
Intro finished, looking for fig and caption of eigenvectors

As we mentioned in Section \ref{subsec:bayesian inverse problems},
$\fwd \prcov \fwd^*$ is the prior covariance $\hilo$ and could be
safely assumed invertible. Proof of Proposition \ref{prop:twice
  woodbury} is delegated to the Supplementary. It amounts to using
Woodbury's matrix identity twice. The standard proof for Woodbury's
matrix identity works for separable Hilbert spaces, as long as all
terms are well defined. Unfortunately, $\obs^*\obs$ is not invertible,
so we utilize a regularization trick to force it to be.

@@

The proof of Proposition \ref{prop:kkt} amounts to utilizing the
Karush-Kuhn-Tucker conditions and is delegated to the
Supplementary. The final ingredient we require for the proof of
Theorem \ref{thm:char} is:

@@

The proof of Lemma is delegated to the Supplementary. It is however
important to note that this proof is construcive. It will give us a
way to construct D-optimal designs, once we fully characterize them in
Theorem \ref{thm:char} below.

@@

  Then:
  \begin{enumerate}
  \item  \(\tr{\opt^*\opt} = m\).
  \item \(\opt^*\opt\) and \(\fwd\prcov\fwd^*\) are simultaneously
    diagonalizable.
  \item \(k := \rank \opt^*\opt \leq m\) and
    \begin{equation*}
      \tar(\opt) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). %= \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
    \end{equation*}
  %% \item 
  %%   \begin{equation*}
  %%     k = \argmax \left \{ k:\lambda_k^{-1} < \sigma^{-2}\frac{m}{k} + \frac{1}{k} \sum_{j=1}^{k}
  %%     \lambda_j^{-1} \right \}.
  %%   \end{equation*}
  \item
    \begin{equation*}
        \eta_i = \begin{cases}
          \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & 1 \leq i \leq k \\
          0 & i > k 
        \end{cases}.
    \end{equation*}
  \item The covariance of the pushforwad \(\fwd_{*} \postopt\) is \(\left
    ( (\fwd \prcov \fwd^*)^{-1} + \sigma^{-2} \opt^*\opt \right
    )^{-1}\) and its eigenvalues are
    \begin{equation*}
      \theta_i =
      \begin{cases}
        \left(\frac{\sum_{j=1}^k \lambda_j^{-1} + \sigma^{-2}m}{k} \right )^{-1} & i \leq k \\
        \lambda_i &  i > k 
      \end{cases}
    \end{equation*}
  \end{enumerate}
\end{theorem}

@@

  Recall from \eqref{eq:postcov}, that the posterior precision is
  $\postcov^{-1} = \prcov^{-1} + \sigma^{-2}\fwd^*\obs^*\obs\fwd$. The
  first statement in part (5) now follows from Proposition
  \ref{prop:twice woodbury}, while the second statement follows from
  parts (1) and (4).
\end{proof}

@@

Theorem \ref{thm:char} facilitates understanding of D-optimal designs
when $\modcov = 0$: Imagine each eigenvector of $\fwd \prcov \fwd^*$
corresponds to a graduated lab cylinder.
%% (Figure \ref{fig:cylinder})
Each cylinder $i$ is filled, a-priori, with green liquid of
$\lambda_i^{-1}$ volume units. We are allowed $m$ measurement, so we
have blue liquid of volume $\sigma^{-2}m$ units at our disposal. When
we seek a D-optimal design, we distribute the blue liquid by
repeatedly adding a drop to whatever cylinder currently has the lowest
level of liquid in it, as long as its index $i \leq m$. The result of
such a procedure is illustrated in Fig.~\ref{fig:optimal vs not}.
%% \begin{figure}%{r}{0.25\textwidth} 
%%     \centering
%%     \includegraphics[width=5cm, height=5cm]{cylinders.jpg}
%%     \caption{Three graduated lab cylinders, corresponding to three
%%       eigenvectors. Prior eigenvalues not shown.}
%%     \label{fig:cylinder}
%% \end{figure}

>>>>

As we mentioned in Section \ref{subsec:bayesian inverse problems},
$\fwd \prcov \fwd^*$ is the prior covariance $\hilo$ and could be
safely assumed invertible. The proof of Proposition \ref{prop:twice
  woodbury} is delegated to the Supplementary. It amounts to using
Woodbury's matrix identity twice. The standard proof for Woodbury's
matrix identity works for separable Hilbert spaces, as long as all
terms are well defined. Unfortunately, $\obs^*\obs$ is not invertible,
so we utilize a regularization trick to force it to be.

@@

The proof of Proposition \ref{prop:kkt} amounts to utilizing the
Karush-Kuhn-Tucker conditions and is delegated to the
Supplementary. The final ingredient we require for the proof of
Theorem \ref{thm:char} characterizing D-optimal designs is:

@@

The proof of Lemma is also delegated to the Supplementary. It is
however important to note that this proof is construcive. It will give
us a way to construct D-optimal designs, once we fully characterize
them in Theorem \ref{thm:char} below.

@@

  Then:
  \begin{enumerate}
  \item  \(\tr{\opt^*\opt} = m\).
  \item \(\opt^*\opt\) and \(\fwd\prcov\fwd^*\) are simultaneously
    diagonalizable.
  \item \(k := \rank \opt^*\opt \leq m\) and
    \begin{equation*}
      \tar(\opt) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). %= \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
    \end{equation*}
  %% \item 
  %%   \begin{equation*}
  %%     k = \argmax \left \{ k:\lambda_k^{-1} < \sigma^{-2}\frac{m}{k} + \frac{1}{k} \sum_{j=1}^{k}
  %%     \lambda_j^{-1} \right \}.
  %%   \end{equation*}
  \item
    \begin{equation*}
        \eta_i = \begin{cases}
          \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & 1 \leq i \leq k \\
          0 & i > k 
        \end{cases}.
    \end{equation*}
  \item The covariance of the pushforwad \(\fwd_{*} \postopt\) is \(\left
    ( (\fwd \prcov \fwd^*)^{-1} + \sigma^{-2} \opt^*\opt \right
    )^{-1}\) and its eigenvalues are
    \begin{equation}\label{eq:cylinders}
      \theta_i =
      \begin{cases}
        \left(\frac{\sum_{j=1}^k \lambda_j^{-1} + \sigma^{-2}m}{k} \right )^{-1} & i \leq k \\
        \lambda_i &  i > k 
      \end{cases}
    \end{equation}
  \end{enumerate}
\end{theorem}

@@

  Recall from \eqref{eq:postcov}, that the posterior precision is
  $\postcov^{-1} = \prcov^{-1} + \sigma^{-2}\fwd^*\opt^*\opt\fwd$. The
  first statement in part (5) now follows from Proposition
  \ref{prop:twice woodbury}, while the second statement follows from
  parts (1) and (4).
\end{proof}

@@

Part (5) of Theorem \ref{thm:char} gives us deep understanding of
D-optimal designs when $\modcov = 0$: Imagine each eigenvector of the
\emph{precision} $\left(\fwd \prcov \fwd^*\right )^{-1}$ corresponds
to a graduated lab cylinder. % (see Fig\ref{fig:cylinder}) Each
cylinder $i$ is filled, a-priori, with green liquid of
$\lambda_i^{-1}$ volume units. We are allowed $m$ measurement, so we
have blue liquid of volume $\sigma^{-2}m$ units at our disposal. When
we seek a D-optimal design, we distribute the blue liquid by
repeatedly adding a drop to whatever cylinder currently has the lowest
level of liquid in it, as long as its index $i \leq m$. The result of
such a procedure is that the precision for the first $k$ eigenvectors
is the average of the total precision $\sum_{j=1}^k \lambda_j^{-1} +
\sigma^{-2}m$, see eq.~\eqref{eq:cylinders} and Fig.~\ref{fig:optimal
  vs not} for an illustration.
%% \begin{figure}%{r}{0.25\textwidth} 
%%     \centering
%%     \includegraphics[width=5cm, height=5cm]{cylinders.jpg}
%%     \caption{Three graduated lab cylinders, corresponding to three
%%      eigenvectors. Prior precision eigenvalues not shown.}
%%     \label{fig:cylinder}
%% \end{figure}

\subsection{Answer to Question \ref{q:why}}\label{subsec:why}
Theorem \ref{thm:char} helps us give a compelling explanation of the
measurement clusterization we obeserved for the inverse problem of the
heat equation (Fig.~\ref{fig:}). Below, we take the following steps:
We first utilize Theorem \ref{thm:char} to understand what a D-optimal
design will look like for our relaxed model; we then argue that
although not identical, a D-optimal design for the full inverse
problem of the heat equation shares similar charachteristics to the
relaxed design; finally, we show how ...

First, recall the relevant parts of Theorem \ref{thm:char}. In our
relaxed model, for a D-optimal design $\opt$:
\begin{itemize}
  \item $\opt^*\opt$ and $\fwd\prcov \fwd^*$ are simultaneously
    diagonalizable.
  \item $\opt^*\opt$ has non-zero eigenvalues only for the $k:=\rank
    \opt^*\opt$ largest eigenvalues $\lambda_j$ of $\fwd\prcov\fwd^*$,
    i.e.~$\eta_j=0$ for $j>k$.
  \item $k$ is monotonic in $m$, the number of measurements we take.
\end{itemize}

Consider $\fwd$ and $\prcov$ from the \emph{the inverse problem of the
heat equation}. As before, we denote the eigenvalues of
$\fwd\prcov\fwd^*$ by $\lambda_j$. We input these eigenvalues into our
\emph{relaxed} model, and find a D-optimal design $\opt$ for our
relaxed model, e.g.~by using Lemma \ref{lemma:free} in the proof of
Theorem \ref{thm:char}. In our relaxed model, the measurements we take
are best utilized in reducing uncertainty for the first $k$
eigenvectors. So, a D-optimal design arising from our \emph{relaxed
model} completely avoids measuring eigenvectors $k+1$ and above.

Of course, in a real life problem --- such as the inverse problem of
the heat equation --- it is likely impossible to find measurement
locations for which all eigenvectors $k+1$ and above are
zero. Luckily, the eigenvalues of $\fwd\prcov\fwd^*$ decay very
quickly: for example, recall from Section \ref{subsec:toy} that the
eigenvalues of the forward model for the 1D heat equation are $-(2\pi
n)^{-2}$. Thus a D-optimal design is a trade-off between measuring the
small number of eigenvectors we are interested in.

We explore the abovementione trade-off for the inverse problem of the
heat equation in Fig.~\ref{fig:eigenvectors}. We allow $m=4$
measurements in $\Omega = [0,1]$ and observe that D-optimal
measurement locations are clustered at $x_1 = 0.31$ and $x_2 =
0.69$. Upon close inspection of the scaled eigenvectors of $\fwd
\prcov \fwd^*$, we first observe that eigenvectors $3$ and above have
negligible prior amplitude. Since we only have $m=4$ measurements at
our disposal, we interpret these results, following Theorem
\ref{thm:char}, as implying we should only care about measuring the
first and second eigenvectors. Then, we note the D-optimal $x_1,x_2$
present a compromise between the amplitude of the first and second
eigenvectors. For example, a measurement at $x=0.5$ would have ignored
the second eigenvector altogether, since the second eigenvector is
zero at $x=0.5$.  Now we can understand measurement clusterization:
attempting to measure what we want and ignoring the rest, along with
the spatial limitations on measurements imposed by the structure of
eigenvectors of $\fwd \prcov \fwd^*$ is the cause for measurement
clusterization in D-optimal designs: we have $m=4$ measuremnts at our
disposal but only two spatial locations that are a good compromise
between the scaled eigenvectors. Thus, clusterization arises by the
pigeonhole principle.

### 2024-07-24 latex/why.tex
Intro finished, looking for fig and caption of eigenvectors

\section{Answer to Question \ref{q:why}: The cause of clusterization}\label{section:why}
According to Theorem \ref{thm:char}, D-optimal designs aim to capture
a small subset of eigenvectors of the prior covariance, specifically
the $k$ eigenvectors with the highest prior variance. Our model
naturally achieves this objective by not measuring eigenvectors $k+1$
and above, as proven in Theorem \ref{thm:char}. Translating this
understanding to spatial problems, we anticipate that a D-optimal
design would favor measurement locations where eigenvectors $k+1$ and
above are either close to zero in value or possess small eigenvalues
in the prior spectrum, for some $k > 0$. To illustrate this
preference, Fig.~\ref{fig:why} depicts the scenario using the 1D
heat equation with homogeneous Dirichlet boundary conditions (details
in the supplementary material). The plot showcases four eigenvectors,
scaled according to their prior standard deviations. Since
eigenvectors beyond the fourth have insignificant prior eigenvalues,
we exclude them from consideration. Notably, we observe that
measurements are clustered near the zeros of the third and fourth
eigenvectors, so we conclude that $k=2$. Clusterization arises because
there are only two locations where the third and fourth eigenvectors
approach zero, while the first and second eigenvectors exhibit
significantly non-zero values. Consequently, when we allocate four
measurements to these two locations, they naturally cluster together,
aligning with the pigeonhole principle.

>>>>

\section{Answer to Question \ref{q:why}: The cause of clusterization}\label{section:why}

%% In the inverse problem of the heat equation, \ev_3, \ev_4,\dots$
%% implies we seek measurement locations for which $\lambda_1\ev_1,
%% \lambda_2\ev_2$ are large, while $\lambda_3\ev_3,
%% \lambda_3\ev_3,\dots$ are small. It turns out that there are two
%% locations that give the best trade-off, and measurements are taken in
%% those two locations. Once we have more than two measurements at our
%% disposal, measurement locations becomde clustered, by the pigeonhole
%% principle, see Fig.~\ref{fig:eigenvectors}.

%% The way a D-optimal design tries to ignore all eigenvectors
%% corresponding to $j > 2$. Since $\lambda_j$ decay to zero, a D-optimal
%% design should only try to avoid eigenvectors for which $\lambda_j$ is
%% not already small --- e.g.~eigenvectors $j=3,4$. It is possible that
%% the geometry of the domain $\Omega$ and the structure of the
%% eigenvectors leave only a few locations in $\Omega$ where the
%% amplitude of eigenvectors $j=1,2$ is high while the amplitude of
%% eigenvectors $j=3,4$ is low. In such case, clusterization will occur
%% whenever the number of measurements $m$ is large enough, by the
%% pigeonhole principle. This scenario is illustrated for the inverse
%% problem of the 1D heat equation in Fig.~\ref{fig:eigenvectors}.

\begin{figure}\label{fig:eigenvectors}
    \centering
    \includegraphics[width=\textwidth]{eigenvectors_dst_scaled.png}
    \caption{D-optimal measurement locations ($m=4$ measurements) and
      weighted eigenvectors for finding the initial condition of
      the 1D heat equation. Measurement locations and weighted
      eigenvectors are plotted over the computational domain $\Omega =
      [0, 1]$ (x-axis). Measurement clusterization occurs
      approximately at $0.31$ and $0.69$. These two locations are a
      compromise between zeros of eigenvectors a D-optimal design aims
      to ignore (third and up) and staying far from zeros of the
      eigenvectors a D-optimal design aims to measure (first and
      second). Allocating $m=4$ measurements into two locations
      results in clusterization, according to the pigeonhole
      principle.}
  \label{fig:why}
\end{figure}

%% According to Theorem \ref{thm:char}, D-optimal designs aim to capture
%% a small subset of eigenvectors of the prior covariance, specifically
%% the $k$ eigenvectors with the highest prior variance. Our model
%% naturally achieves this objective by not measuring eigenvectors $k+1$
%% and above, as proven in Theorem \ref{thm:char}. Translating this
%% understanding to spatial problems, we anticipate that a D-optimal
%% design would favor measurement locations where eigenvectors $k+1$ and
%% above are either close to zero in value or possess small eigenvalues
%% in the prior spectrum, for some $k > 0$. To illustrate this
%% preference, Fig.~\ref{fig:why} depicts the scenario using the 1D
%% heat equation with homogeneous Dirichlet boundary conditions (details
%% in the supplementary material). The plot showcases four eigenvectors,
%% scaled according to their prior standard deviations. Since
%% eigenvectors beyond the fourth have insignificant prior eigenvalues,
%% we exclude them from consideration. Notably, we observe that
%% measurements are clustered near the zeros of the third and fourth
%% eigenvectors, so we conclude that $k=2$. Clusterization arises because
%% there are only two locations where the third and fourth eigenvectors
%% approach zero, while the first and second eigenvectors exhibit
%% significantly non-zero values. Consequently, when we allocate four
%% measurements to these two locations, they naturally cluster together,
%% aligning with the pigeonhole principle.
